### 2010-04-01 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: align codes, whenever possible
Desc@ICFP: fix colors
Desc@ICFP: proofread section 5, 'Indexed description'

Evolving in a dependent setting fosters new opportunities for
datatypes. The typical example is bounded lists, also known as
vectors. A vector is a list decorated by its length. Having this
information prevents hazardous operations, such as taking the head of
an empty vector: the head function only takes vectors of length
$\NatSuc{\V{n}}$, as enforced by its type. This is made possible by
the specificity of dependent types: a term -- the length -- can
influence a type -- the vector type.

However, these datatypes cannot be defined by mere induction. In the
case of vectors, for instance, we have to define the whole
\emph{family} of vectors in one go: vectors of all sizes need to be
defined at the same time. In dependently-typed languages, the basic
grammar of datatypes is that of inductive families. To capture this
grammar, we rely on \emph{indexing}. We are going to transform the
$\Desc$ universe into $\IDesc{\!}$, the universe of indexed
descriptions. We model inductive families by the $\V{I} \To
\IDesc{\V{I}}$ type.

@@

In the previous section, we have presented the $\Desc$ universe as the
grammar of functors in the category $\Set$. We have seen how to code
inductive types in this setting. To describe an inductive family
indexed by $\Bhab{\V{I}}{\Set}$, we use endofunctors on the category
$\Set^{\V{I}}$. We call these \emph{indexed functors}. \note{Predicate
  transformer anyone?} $\V{I} \To \IDesc{\V{I}}$ is our
grammar for describing these functors. Hence, $\IDesc{\!}$ and its
interpretation have the following types:

@@

However, we still have to define the actual grammar. We obtain it by
evolving $\Desc$ to cope with indexing. \note{Say something about the
  transformation} The code of $\IDesc{\!}$ is presented in
Figure~\ref{fig:idesc}. The induction scheme on $\IDesc{\!}$ has the
following type:

\[\stk{
\begin{array}{@{}ll}
\F{inductionI} : & _{\Bhab{\V{I}}{\Set}}
                   \PITEL{\V{R}}{\V{I} \To \IDesc{\V{I}}}
                   \PI{\V{P}}{\SIGMA{\V{i}}{\V{I}} \IMu{\V{I}}{\V{R}}{\!} \To \Set} \\
                 & (      \PITEL{\V{i}}{\V{I}} 
                          \PI{\V{xs}}{\idescop{\V{R}\: \V{i}}{\V{I}}{\IMu{\V{I}}{\V{R}}{\!}}} \\
                 & \   \idescop{\AllI{\V{I}}
                                     {(\V{R}\: \V{i})}
                                     {(\IMu{\V{I}}{\V{R}}{\!})}
                                     {\V{xs}}}
                               {}
                               {\V{P}} \To
                       \V{P}\: \pair{\V{i}}{\Con{\V{xs}}}{}) \\
                 & \PITEL{\V{i}}{\V{I}}
                   \PI{\V{x}}{\IMu{\V{I}}{\V{R}}{\V{i}}}
                   \V{P}\: \pair{\V{i}}{\V{x}}{}
\end{array} \\
\F{inductionI}\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: (\Con{\V{xs}}) =  \\
\qquad
    \V{ms}\: \V{i}\: \V{xs}\: (\allI{}
                                    {\V{R}\: \V{i}}
                                    {\IMu{\V{I}}{R}{}}
                                    {\V{P}}
                                    {(\LAM{\pair{\V{i}}{\V{xs}}{}} \F{inductionI}\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: \V{xs})}
                                    {\V{xs}})
}\]

Where the operators $\AllI{\!}{\!}{\!}{\!}$ and
$\allI{\!}{\!}{\!}{\!}{\!}{\!}$ are defined in
Figure~\ref{fig:allI-predicates}. As for descriptions, we can compute
a generic catamorphism, $\F{cataI}$, from $\F{inductionI}$.

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\AllI{\!}{\!}{\!}{\!} : & _{\Bhab{\V{I}}{\Set}}
                          \PITEL{\V{D}}{\IDesc{\V{I}}}
                          \PI{\V{X}}{\V{I} \To \Set} \\
                        & \idescop{\V{D}}{\V{I}}{\V{X}} \To
                          \IDesc{(\TIMES{\V{I}}{\V{X}})}
\end{array} \\
\begin{array}{@{}ll@{\:=\:\:}l}
\AllI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{x}} &
    \DVar{\pair{\V{i}}{\V{x}}{}} \\
\AllI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{k}} &
    \DConst{K} \\
\AllI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\pair{\V{d}}{\V{d'}}{}} &
    \DProd{\AllI{}{\V{D}}{\V{X}}{\V{d}}}{\AllI{}{\V{D'}}{\V{X}}{\V{d'}}} \\
\AllI{\:}{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\pair{\V{a}}{\V{b}}{}} &
    \AllI{}{(\V{T}\: \V{a})}{\V{X}}{\V{b}} \\
\AllI{\:}{(\DPi{\V{S}}{\V{T}})}{& \V{X}}{\V{f}} &
    \DPi{\V{S}}{(\LAM{s} \AllI{}{(\V{T}\: \V{s})}{\V{X}}{(\V{f}\: \V{s})})}
\end{array}
}
&
%%
\stk{
\begin{array}{@{}ll}
\allI{\!}{\!}{\!}{\!}{\!}{\!} : & _{\Bhab{\V{I}}{\Set}}
                                  \PITEL{\V{D}}{\IDesc{\V{I}}}
                                  \PITEL{\V{X}}{\V{I} \To \Set} 
                                  \PI{\V{R}}{\TIMES{\V{I}}{\V{X}} \To \Set} \\
                                & (\PI{\V{x}}{\TIMES{\V{I}}{\V{X}}} \V{R}\: \V{x}) \To
                                  \PI{\V{xs}}{\idescop{\V{D}}{}{\V{X}}} 
                                  \idescop{\AllI{}{\V{D}}{\V{X}}{\V{xs}}}{\TIMES{\V{I}}{\V{X}}}{\V{R}}
\end{array} \\
\begin{array}{@{}ll@{\:=\:\:}l}
\allI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{R}}{\V{P}}{\V{x}} &
    \V{P}\: \pair{\V{i}}{\V{x}}{} \\
\allI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{R}}{\V{P}}{\V{k}} &
    \V{k} \\
\allI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\V{R}}{\V{P}}{(\V{d}, \V{d'})} &
    \pair{\allI{}{\V{D}}{\V{X}}{\V{R}}{\V{P}}{\V{d}}}
         {\allI{}{\V{D'}}{\V{X}}{\V{R}}{\V{P}}{\V{d'}}}{} \\
\allI{\:}{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\V{R}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} &
    \allI{}{(\V{T}\: \V{a})}{\V{X}}{\V{R}}{\V{P}}{\V{b}} \\
\allI{\:}{(\DPi{\V{S}}{\V{T}})}{& \V{X}}{\V{R}}{\V{P}}{\V{f}} &
    \allI{}{(\V{T}\: \V{a})}{\V{X}}{\V{R}}{\V{P}}{(\V{f}\: \V{a})}
\end{array}
\end{array}
}
\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \IDesc{(\Bhab{\V{I}}{\Set})} : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{\!}         & \V{I} \To \IDesc{\V{I}}                                       \\
    \DConst{\!}       & \Set \To \IDesc{\V{I}}                                        \\
    \DProd{\!}{\!}    & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}       \\
    \DSigma{\!}{\!}   & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
    \DPi{\!}{\!}      & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}~\V{i}                                                 \\
\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
\end{array}
}
\end{array}
}\]

@@

Because $\Nat$ is just an inductive type, $\NatD$ is a $\Unit$-indexed
functor. Therefore, the recursive argument is materialised by
$\DVar{\Void}$, where we were using $\DIndx$ in the previous
presentation. This transformation generalizes to all inductive
types. Moreover, we gain the ability to write mutually recursive
inductive types.

@@

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescD~\V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \red{[} 
                                          \DVar{\!},
                                          \DConst{\!},
                                          \DProd{\!}{\!},
                                          \DSigma{\!}{\!}, 
                                          \DPi{\!}{\!}
                                             \red{]}) \\
                              & \bigRedBracket{\begin{array}{l}
                                      \DConst{\V{I}}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             }
\end{array}
}\]

@@

So far, the examples we have seen lives in $\IDesc{\Unit}$, hence are
not using any indexing. We remedy this by encoding the vectors. Recall
that the constructors $\VecNil$ and $\VecCons$ are only defined for an
index $\NatZero$ and $\NatSuc$ respectively.:

\[
\stk{
\data \D{Vec}\: \PITEL{\V{X}}{\Set} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \VecNil      & \D{Vec}\:\V{X}\:{\NatZero)}   \\
    \VecCons     & _{\PI{\V{n}}{\Nat}}\V{X} \To \D{Vec}\:{\V{X}}\:{\V{n}} \To \D{Vec}\:{\V{X}}\:{(\NatSuc{\V{n}})}
\end{array}
}
\]

One way to code constrained datatype is to appeal to equality. The
constraints are therefore captured by equations in the datatype. In
this case, we obtain the following definition:

@@

In the $\VecNil$ case, a proof must be provided that the index is
equal to $\NatZero$. In the $\VecCons$ case, we first store an element
$\V{m}$ of $\Nat$. However, the constraint stipulates that $\V{m}$
cannot be \emph{any} natural numbers: it must be ``the index minus
one''. This translates into the constraint $\V{n} \PropEq
\NatSuc{\V{m}}$.

However, one could, quite rightfully, be worried by the introduction
of equalities in the picture. In a programming context, equalities
haunt us, begging for proofs. Hence, whereas the developer typed a
constructor-constrained datatype, the elimination form will involve
equalities and their proofs. As much as possible, one would like to
avoid suffering from this burden. However, the equalities introduced
here are rather special: they are \emph{equations} constraining
variables to certain terms. \citet{goguen:pattern-matching} have
shown, in the context of dependent pattern matching, how to tame these
equations.

@@

\paragraph{Tagged Indexed Descriptions:}

@@

For convenience, we divide a tagged indexed description in two parts:
first, the constructors that do not depend on the index; then, the
constructors that do. The non-dependent part mirrors the definition
for non-indexed descriptions. The index-depend part simply indexes the
choice of constructors by $\V{I}$. Hence, by inspecting the index, it
is possible to enable or disable the constructors.

\[\stk{
 \TagIDesc{\V{I}}  \mapsto \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
 \ATagIDesc{\V{I}} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
 \ITagIDesc{\V{I}} \mapsto \\
\qquad\SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
}\]

@@

%% Constructors
\newcommand{\Eval}[1]{\etag{\CN{val}}~#1}
\newcommand{\Econd}[3]{\etag{\CN{cond}}~#1~#2~#3}
\newcommand{\Eplus}[2]{\etag{\CN{plus}}~#1~#2}
\newcommand{\Ele}[2]{\etag{\CN{le}}~#1~#2}

%% Index mapper (terminology?)
\newcommand{\Val}[1]{\F{Val}~#1}
\newcommand{\Var}[2]{\F{Var}_{#1}~#2}

@@

\[
\begin{array}{l@{\::\:\:}l}
\Econd{}{}{}     & \forall \Bhab{\V{ty}}{\Ty} . \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
\Eplus{}{}       & \Enat \To \Enat \To \Enat                           \\
\Ele{}{}         & \Enat \To \Enat \To \Ebool                          \\
\Eval{}          & \forall \Bhab{\V{ty}}{\Ty} . \Val{\V{ty}} \To \V{ty}
\end{array}
\]

The function $\Val{\!}$, used in the definition of $\Eval{\!}$, simply
maps a type $ty$ of the object language to the corresponding type in
the host language. Hence, the arguments of $\Eval{}$ are ensured to be
of the expected type. We assume that $\Nat$ and $\Bool$ represent
natural numbers and booleans in the host language, equipped with an
addition operation $\F{plusHost}$ and a comparison function
$\F{leHost}$. We define $\Val{}$ as follows:

\[\stk{
\Val{} : \Ty \To \Set \\
\begin{array}{@{}l@{\:=\:\:}l}
\Val{\Enat}   & \Nat \\
\Val{\Ebool}  & \Bool
\end{array}
}\]

In our universe of descriptions, this datatype is represented by a
tagged indexed description. We use the index to carry the type: the
resulting description is indexed by $\Ty$. We observe that some
constructors are always defined, namely $\Econd{}{}{}$ and $\Eval{}$. On the
other hand, the $\Eplus$ and $\Ele$ constructors are
index-dependent. $\Eplus$ is defined if and only if the result type --
the index -- is $\Enat$, whereas $\Ele$ is defined if and only if the
index is $\Ebool$. The actual code precisely follows this intuition,
as shown in Figure~\ref{fig:hexpr-full}. For brevity, we use an
informal $\case{\ldots}{\ldots}$ notation, simulating a definition by
pattern-matching. Formally, this corresponds to a call to the
$\switch{\!}{\!}{\!}{\!}$ eliminator.

@@

\[\stk{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \bigRedBracket{
                 \begin{array}{l}
                   {[\Eval{}, \Econd{\!}{\!}{\!}\!\,]} \red{,} \\
                   \LAM{\V{ty}}
                   \bigRedBracket{
                   \begin{array}{l}
                   \DConst{(\Val{\V{ty}})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
                   \end{array}
                   }
                 \end{array}
                 }
\\
\\
} 
\\
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \LAM{\V{ty}} \case{\V{ty}}{\Enat \To {[\Eplus{\!}{\!}\!\,]} \\ \Ebool \To {[\Ele{\!}{\!}\!\,]}} \red{,} \\
                   \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
                   \end{array}
                   }
}
}\]

@@

However, so far, we are only able to define and manipulate
\emph{closed} terms. By abstracting over $\Val{}$, it is possible to
build and manipulate \emph{open} terms, that is, terms with
variables. On the model of $\Val{}$, we define $\Var{}{}$:

>>>>

Working with dependent types fosters new opportunities for
datatypes. The typical example is bounded lists, also known as
vectors. A vector is a list decorated by its length. Having this
information prevents hazardous operations, such as taking the head of
an empty vector: the head function only takes vectors of length
$\NatSuc{\V{n}}$, as enforced by its type. This is made possible by
the specificity of dependent types: a term -- the length -- can
influence a type -- the vector type.

However, these datatypes cannot be defined by mere induction. In the
case of vectors, for instance, we have to define the whole
\emph{family} of vectors in one go: vectors of all sizes need to be
defined at the same time. In dependently-typed languages, the basic
grammar of datatypes is that of inductive families. To capture this
grammar, we rely on \emph{indexing}.

@@

In the previous section, we have presented the $\Desc$ universe as a
grammar of functors in the category $\Set$. We have seen how to code
inductive types in this setting. To describe an inductive family
indexed by $\Bhab{\V{I}}{\Set}$, we use endofunctors on the category
$\Set^{\V{I}}$. We call these \emph{indexed functors}. \note{Predicate
  transformer anyone?} $\V{I} \To \IDesc{\V{I}}$ is our grammar for
describing these functors. Hence, $\IDesc{\!}$ and its interpretation
have the following types:

@@

However, we still have to define the actual grammar. We obtain it by
evolving $\Desc$ to cope with indexing. \note{Say something about the
  transformation} The code of $\IDesc{\!}$ is presented in
Figure~\ref{fig:idesc}. Induction on indexed descriptions is defined
by:

\[\stk{
\begin{array}{@{}ll}
\F{inductionI} : & _{\PI{\V{I}}{\Set}}
                   \PITEL{\V{R}}{\V{I} \To \IDesc{\V{I}}}
                   \PI{\V{P}}{\SIGMA{\V{i}}{\V{I}} \IMu{\V{I}}{\V{R}}{\!} \To \Set} \\
                 & (      \PITEL{\V{i}}{\V{I}} 
                          \PI{\V{xs}}{\idescop{\V{R}\: \V{i}}{\V{I}}{\IMu{\V{I}}{\V{R}}{\!}}} \\
                 & \   \idescop{\AllI{\V{I}}
                                     {(\V{R}\: \V{i})}
                                     {(\IMu{\V{I}}{\V{R}}{\!})}
                                     {\V{xs}}}
                               {}
                               {\V{P}} \To
                       \V{P}\: \pair{\V{i}}{\Con{\V{xs}}}{}) \\
                 & \PITEL{\V{i}}{\V{I}}
                   \PI{\V{x}}{\IMu{\V{I}}{\V{R}}{\V{i}}}
                   \V{P}\: \pair{\V{i}}{\V{x}}{}
\end{array} \\
\F{inductionI}\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: (\Con{\V{xs}}) =  \\
\quad
    \V{ms}\: \V{i}\: \V{xs}\: (\allI{}
                                    {\V{R}\: \V{i}}
                                    {\IMu{\V{I}}{R}{}}
                                    {\V{P}}
                                    {(\LAM{\pair{\V{i}}{\V{xs}}{}} \F{inductionI}\: \V{R}\: \V{P}\: \V{ms}\: \V{i}\: \V{xs})}
                                    {\V{xs}})
}\]

Where the operators $\AllI{}{\!}{\!}{}$ and
$\allI{}{\!}{\!}{\!}{\!}{}$ are presented in
Figure~\ref{fig:allI-predicates}. As for descriptions, we can compute
a generic catamorphism, $\F{cataI}$, from $\F{inductionI}$.

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\AllI{\!}{\!}{\!}{\!} : & _{\PI{\V{I}}{\Set}}
                          \PITEL{\V{D}}{\IDesc{\V{I}}}
                          \PI{\V{X}}{\V{I} \To \Set} \\
                        & \idescop{\V{D}}{\V{I}}{\V{X}} \To
                          \IDesc{(\TIMES{\V{I}}{\V{X}})}
\end{array} \\
\begin{array}{@{}l@{}l@{\:=\:\:}l}
\AllI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{x}} &
    \DVar{\pair{\V{i}}{\V{x}}{}} \\
\AllI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{k}} &
    \DConst{\V{K}} \\
\AllI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\pair{\V{d}}{\V{d'}}{}} &
    \DProd{\AllI{}{\V{D}}{\V{X}}{\V{d}}}{\AllI{}{\V{D'}}{\V{X}}{\V{d'}}} \\
\AllI{\:}{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\pair{\V{a}}{\V{b}}{}} &
    \AllI{}{(\V{T}\: \V{a})}{\V{X}}{\V{b}} \\
\AllI{\:}{(\DPi{\V{S}}{\V{T}})}{& \V{X}}{\V{f}} &
    \DPi{\V{S}}{(\LAM{\V{s}} \AllI{}{(\V{T}\: \V{s})}{\V{X}}{(\V{f}\: \V{s})})}
\end{array}
}
&
%%
\stk{
\begin{array}{@{}ll}
\allI{\!}{\!}{\!}{\!}{\!}{\!} : & _{\PI{\V{I}}{\Set}}
                                  \PITEL{\V{D}}{\IDesc{\V{I}}}
                                  \PITEL{\V{X}}{\V{I} \To \Set} 
                                  \PI{\V{R}}{\TIMES{\V{I}}{\V{X}} \To \Set} \\
                                & (\PI{\V{x}}{\TIMES{\V{I}}{\V{X}}} \V{R}\: \V{x}) \To
                                  \PI{\V{xs}}{\idescop{\V{D}}{}{\V{X}}} 
                                  \idescop{\AllI{}{\V{D}}{\V{X}}{\V{xs}}}{\TIMES{\V{I}}{\V{X}}}{\V{R}}
\end{array} \\
\begin{array}{@{}l@{}l@{\:=\:\:}l}
\allI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{R}}{\V{P}}{\V{x}} &
    \V{P}\: \pair{\V{i}}{\V{x}}{} \\
\allI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{R}}{\V{P}}{\V{k}} &
    \V{k} \\
\allI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\V{R}}{\V{P}}{(\V{d}, \V{d'})} &
    \pair{\allI{}{\V{D}}{\V{X}}{\V{R}}{\V{P}}{\V{d}}}
         {\allI{}{\V{D'}}{\V{X}}{\V{R}}{\V{P}}{\V{d'}}}{} \\
\allI{\:}{(\DSigma{\V{S}}{\V{T}})}{& \V{X}}{\V{R}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} &
    \allI{}{(\V{T}\: \V{a})}{\V{X}}{\V{R}}{\V{P}}{\V{b}} \\
\allI{\:}{(\DPi{\V{S}}{\V{T}})}{& \V{X}}{\V{R}}{\V{P}}{\V{f}} &
    \allI{}{(\V{T}\: \V{a})}{\V{X}}{\V{R}}{\V{P}}{(\V{f}\: \V{a})}
\end{array}
\end{array}
}
\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \IDesc{(\Bhab{\V{I}}{\Set})} : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{\!}         & \V{I} \To \IDesc{\V{I}}                                       \\
    \DConst{\!}       & \Set \To \IDesc{\V{I}}                                        \\
    \DProd{\!}{\!}    & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}       \\
    \DSigma{\!}{\!}   & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
    \DPi{\!}{\!}      & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}\: \V{i}                                           \\
\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
\end{array}
}
\end{array}
}\]

@@

Because $\Nat$ is just an inductive type, $\NatD$ is a $\Unit$-indexed
functor. Therefore, the recursive argument is materialised by
$\DVar{\Void}$, where we were using $\DIndx{\!}$ in the previous
presentation. This transformation generalizes to all inductive
types. Moreover, we gain the ability to write mutually recursive
inductive types.

@@

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescD\: \V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \red{[} 
                                          \DVar{\!},
                                          \DConst{\!},
                                          \DProd{\!}{\!},
                                          \DSigma{\!}{\!}, 
                                          \DPi{\!}{\!}
                                             \red{]}) \\
                              & \bigRedBracket{\begin{array}{l}
                                      \DConst{\V{I}}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             }
\end{array}
}\]

@@

So far, the examples we have seen lives in $\IDesc{\Unit}$, hence are
not using any indexing. We remedy this by encoding the vectors. Recall
that the constructors $\VecNil$ and $\VecCons{\!}{\!}$ are only defined for an
index $\NatZero$ and $\NatSuc$ respectively.:

\[
\stk{
\data \D{Vec}\: \PITEL{\V{X}}{\Set} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \VecNil          & \D{Vec}\:\V{X}\:{\NatZero)}   \\
    \VecCons{\!}{\!} & _{\PI{\V{n}}{\Nat}}\V{X} \To \D{Vec}\:{\V{X}}\:{\V{n}} \To \D{Vec}\:{\V{X}}\:{(\NatSuc{\V{n}})}
\end{array}
}
\]

One way to code constrained datatypes is to appeal to equality. The
constraints are therefore captured by equations in the datatype. In
this case, we obtain the following definition:

@@

In the $\VecNil$ case, a proof must be provided that the index is
equal to $\NatZero$. In the $\VecCons{\!}{\!}$ case, we first store an element
$\V{m}$ of $\Nat$. However, the constraint stipulates that $\V{m}$
cannot be \emph{any} natural numbers: it must be ``the index minus
one''. This translates into the constraint $\V{n} \PropEq
\NatSuc{\V{m}}$.

However, one could, quite rightfully, be worried by the introduction
of equalities in the picture. In a programming context, equalities
haunt us, begging for proofs. As much as possible, one would like to
avoid suffering from this burden. However, the equalities introduced
here are rather special: they are \emph{equations} constraining
variables to certain terms. \citet{goguen:pattern-matching} have
shown, in the context of dependent pattern matching, how to tame these
equations.

@@

\paragraph{Tagged indexed descriptions:}

@@

We divide a tagged indexed description in two parts: first, the
constructors that do not depend on the index; then, the constructors
that do. The non-dependent part mirrors the definition for non-indexed
descriptions. The index-depend part simply indexes the choice of
constructors by $\V{I}$. Hence, by inspecting the index, it is
possible to enable or disable constructors.

\[
\begin{array}{@{}l@{\:\mapsto\:\:}l}
 \TagIDesc{\V{I}}  & \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
 \ATagIDesc{\V{I}} & \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
 \ITagIDesc{\V{I}} & 
     \SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
\end{array}
\]

@@

%% Constructors
\newcommand{\Eval}[1]{\etag{\CN{val}}\:#1}
\newcommand{\Econd}[3]{\etag{\CN{cond}}\:#1\:#2\:#3}
\newcommand{\Eplus}[2]{\etag{\CN{plus}}\:#1\:#2}
\newcommand{\Ele}[2]{\etag{\CN{le}}\:#1\:#2}

%% Index mapper (terminology?)
\newcommand{\Val}[1]{\F{Val}\:#1}
\newcommand{\Var}[2]{\F{Var}_{#1} #2}

@@

\[
\begin{array}{l@{\::\:\:}l}
\Econd{\!}{\!}{\!}   & \forall \Bhab{\V{ty}}{\Ty} . \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
\Eplus{\!}{\!}       & \Enat \To \Enat \To \Enat                           \\
\Ele{\!}{\!}         & \Enat \To \Enat \To \Ebool                          \\
\Eval{\!}            & \forall \Bhab{\V{ty}}{\Ty} . \Val{\V{ty}} \To \V{ty}
\end{array}
\]

The function $\Val{\!}$, used in the definition of $\Eval{\!}$, simply
maps a type $ty$ of the object language to the corresponding type in
the host language. Hence, the arguments of $\Eval{\!}$ are ensured to
be of the expected type. We assume that $\Nat$ and $\Bool$ represent
natural numbers and booleans in the host language, equipped with an
addition operation $\F{plusHost}$ and a comparison function
$\F{leHost}$. We define $\Val{\!}$ as follows:

\[\stk{
\Val{\!} : \Ty \To \Set \\
\begin{array}{@{}l@{\:=\:\:}l}
\Val{\Enat}   & \Nat \\
\Val{\Ebool}  & \Bool
\end{array}
}\]

In our universe of descriptions, the syntax of this language is
described by a tagged indexed description. We use the index to carry
the type: the resulting description is indexed by $\Ty$. We observe
that some constructors are always defined, namely $\Econd{\!}{\!}{\!}$
and $\Eval{\!}$. On the other hand, the $\Eplus{\!}{\!}$ and
$\Ele{\!}{\!}$ constructors are index-dependent. $\Eplus{\!}{\!}$ is
defined if and only if the result type -- the index -- is $\Enat$,
whereas $\Ele{\!}{\!}$ is defined if and only if the index is
$\Ebool$. The actual code precisely follows this intuition, as shown
in Figure~\ref{fig:hexpr-full}. For brevity, we use an informal
$\case{\ldots}{\ldots}$ notation, simulating a definition by
pattern-matching. Formally, this corresponds to a call to the
$\switch{\!}{\!}{\!}{\!}$ eliminator.

@@

\[\stk{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \bigRedBracket{
                 \begin{array}{l}
                   {[\Eval{\!}, \Econd{\!}{\!}{\!}\!\,]} \red{,} \\
                   \LAM{\V{ty}}
                   \bigRedBracket{
                   \begin{array}{l}
                   \DConst{(\Val{\V{ty}})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
                   \end{array}
                   }
                 \end{array}
                 }
\\
\\
} 
\\
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \LAM{\V{ty}} \case{\V{ty}}{\Enat \To {[\Eplus{\!}{\!}\!\,]} \\ \Ebool \To {[\Ele{\!}{\!}\!\,]}} \red{,} \\
                   \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
                   \end{array}
                   }
}
}\]

@@

However, so far, we are only able to define and manipulate
\emph{closed} terms. By abstracting over $\Val{\!}$, it is possible to
build and manipulate \emph{open} terms, that is, terms with
variables. Following $\Val{\!}$, we define $\Var{}{}$ by:
